Why aren't significance tests commonly used for linear regression diagnostics?

### 2022-11-15 paper_comparison/paper_comparison/paper_comparison.Rmd
finish p-value cmp

                     tibble(shape = shape,
                            e_sigma = e_sigma,
                            x_dist = x_dist,
                            n = n,
                            effect_size = mod$effect_size(tmp_dat),
                            F_p_value = mod$test(tmp_dat)$p_value,
                            BP_p_value = HETER_MODEL$test(tmp_dat)$p_value,
                            SW_p_value = shapiro.test(tmp_dat$.resid)$p.value,
                            boot_id = i)
                   }) %>%
  reduce(bind_rows)
  }

@@

                     tibble(a = a,
                            b = b,
                            x_dist = x_dist,
                            n = n,
                            effect_size = mod$effect_size(tmp_dat, type = "kl"),
                            F_p_value = POLY_MODEL$test(mutate(tmp_dat, z = poly_model()$gen(n, computed = select(tmp_dat, x))$z))$p_value,
                            BP_p_value = mod$test(tmp_dat)$p_value,
                            SW_p_value = shapiro.test(tmp_dat$.resid)$p.value,
                            boot_id = i)
                   }) %>%
  reduce(bind_rows)
  }

@@

## P-value comparison

Visual inference and conventional hypothesis testing are two distinct statistical procedures. As displayed in Figure \ref{fig:p-value-comparison}, the visual test gives higher p-values than the conventional test in both polynomial and heteroskedasticity models in most of the time, suggesting that humans have lower confidence and much higher tolerance of the residual departures than the conventional test.

In particular, given 95% confidence level, both tests will agree on the rejection in only around 58% of the cases. And in 41% of the cases, the visual test does not reject the null hypothesis but the conventional test does. In these 41% of the cases, though we know that the designed lineups contain various degrees of residual departures, analysts and the general public as the consumers of the output may not believe in the rejection of the null hypothesis, or at least may not consider the model violation as impactful, simply due to the fact that the departures are not clearly visible to them.

This makes it sounds like people are arrogantly ignoring the accuracy of conventional tests and trust too much to their visual perception, but it is not. In fact, the sensitivity of the conventional test could sometimes distract and discourage analysts from finding a good linear approximation to the data. The rejection of the null hypothesis because of acceptable and negligible residual departure is not practically useful and thus is usually viewed as "false alarm" by analysts. Relying on visual perception to determine the quality of the regression model is just a way to

Alternatively,

From a practical sense, data plots and the visual test provide better measure of the seriousness of the residual departure.

One noticeable difference compared to the result provided in @buja_statistical_2009 is the lack of data points

Note that

This also partly explains why abundance of literature suggesting the use of data plot in regression diagnostics.

```{r p-value-comparison, fig.height=4, fig.cap="..."}
vi_survey %>%
  filter(!null_lineup) %>%
  filter(!attention_check) %>%
  group_by(unique_lineup_id) %>%
  summarise(across(everything(), first)) %>%
  ggplot() +
  geom_abline() +
  geom_point(aes(conventional_p_value, p_value), alpha = 0.4) +
  scale_x_sqrt() +
  scale_y_sqrt() +
  facet_wrap(~type) +
  ylab("Visual test p-value on square root scale") +
  xlab("Conventional test p-value on square root scale")
```

## Power comparison

Figure \ref{fig:power-poly-uniform} shows the estimated power of visual test on lineups with uniform distribution $X_{raw}$ against natural logarithm of the effect with comparison to the power of an exact test - F-test, and the power of two other residual-based conventional tests commonly used in regression diagnostics but for testing other departures from the model assumptions. In overall, the power of all four tests increases as the effect becomes larger. The power curve of F-test climbs aggressively from 25% to around 90% as $log_e(E)$ increases from 0 to 2, while others respond inactively to the change of effect and remain lower than 25% throughout the period, showing that as an exact test, the F-test is relatively more sensitive to the type of model defects that being considered. The power of visual test arises steadily and nearly linearly to around 90% as $log_e(E)$ increases from 2 to 5, suggesting that the effect starts to make noticeable impact on the degree of the presence of the designed visual features. Other two inappropriate conventional tests shows improvement at the same time but at a lower rate. This coincides the point made by @cook1982residuals mentioned in \ref{hypothesis-testing} that residual-based tests for a specific type of model defect are sensitive to other types of model defects. At $log_e(E) \approx 6$, the power curve of F-test reaches almost 100% followed by the visual test by a small margin. The power of Breusch--Pagan test and Shapiro--Wilk test reach around 75% and 30% respectively.

@@

@@

```{r power-poly-uniform, fig.cap = "..."}
vi_survey %>%
  filter(x_dist == "uniform") %>%
  group_by(unique_lineup_id) %>%
  summarise(across(everything(), first)) %>%
  filter(type == "polynomial") %>%
  filter(e_sigma != 0.125) %>%
  mutate(reject_dirichlet = as.numeric(p_value <= 0.05)) %>%
  mutate(log_effect_size = log(effect_size)) %>%
  ggplot() +
  geom_smooth(aes(log_effect_size, reject_dirichlet, col = "VI_p_value"), method = "glm", method.args = list(family = binomial), se = FALSE) +
  geom_point(aes(log_effect_size, prop_detect), alpha = 0.3) + geom_point(data = ~filter(.x, unique_lineup_id %in% c("poly_502", "poly_78", "poly_6", "poly_447")), aes(log_effect_size, prop_detect), alpha = 1, col = "red", size = 2) +
  stat_smooth(geom = "line", 
              data = ~boot_y(.x, reject_dirichlet, times = 500), 
              aes(log_effect_size, reject_dirichlet, group = boot_id, col = "Visual p-value"), 
              method = "glm", 
              method.args = list(family = binomial), 
              se = FALSE, 
              alpha = 0.03) +
  geom_smooth(data = mutate(poly_conv_sim, log_effect_size = log(effect_size)) %>%
                filter(x_dist == "uniform") %>%
                pivot_longer(F_p_value:SW_p_value), 
              aes(log_effect_size, as.numeric(value < 0.05), col = name), 
              method = "glm", 
              method.args = list(family = binomial),
              se = FALSE) +
  theme_light(base_size = 5) +
  facet_wrap(~n) -> p

VI_MODEL$plot_lineup(vi_lineup[["poly_502"]]$data, 
                     remove_grid_line = TRUE, 
                     theme = theme_light(base_size = 5),
                     remove_axis = TRUE) + 
  ggtitle(bquote(log[e](E) %~~% 1~","~"F-test p-value" == .(format(filter(vi_survey, unique_lineup_id == "poly_502")$conventional_p_value[1], digits = 3))~", Visual p-value " == .(format(filter(vi_survey, unique_lineup_id == "poly_502")$p_value[1], digits = 3)))) -> p1

VI_MODEL$plot_lineup(vi_lineup[["poly_78"]]$data, 
                     remove_grid_line = TRUE, 
                     theme = theme_light(base_size = 5),
                     remove_axis = TRUE) + 
  ggtitle(bquote(log[e](E) %~~% 2~","~"F-test p-value" == .(format(filter(vi_survey, unique_lineup_id == "poly_78")$conventional_p_value[1], digits = 3))~", Visual p-value " == .(format(filter(vi_survey, unique_lineup_id == "poly_78")$p_value[1], digits = 3)))) -> p2

VI_MODEL$plot_lineup(vi_lineup[["poly_6"]]$data, 
                     remove_grid_line = TRUE, 
                     theme = theme_light(base_size = 5),
                     remove_axis = TRUE) + 
  ggtitle(bquote(log[e](E) %~~% 3~","~"F-test p-value" == .(format(filter(vi_survey, unique_lineup_id == "poly_6")$conventional_p_value[1], digits = 3))~", Visual p-value " == .(format(filter(vi_survey, unique_lineup_id == "poly_6")$p_value[1], digits = 3)))) -> p3

VI_MODEL$plot_lineup(vi_lineup[["poly_447"]]$data, 
                     remove_grid_line = TRUE, 
                     theme = theme_light(base_size = 5),
                     remove_axis = TRUE) + 
  ggtitle(bquote(log[e](E) %~~% 4~","~"F-test p-value" == .(format(filter(vi_survey, unique_lineup_id == "poly_447")$conventional_p_value[1], digits = 3))~", Visual p-value " == .(format(filter(vi_survey, unique_lineup_id == "poly_447")$p_value[1], digits = 3)))) -> p4

>>>>

                     es <- filter(vi_survey, 
                                  shape == {{shape}},
                                  e_sigma == {{e_sigma}},
                                  x_dist == {{x_dist}},
                                  n == {{n}}) %>%
                       head(1) %>%
                       pull(effect_size)

                     tibble(shape = shape,
                            e_sigma = e_sigma,
                            x_dist = x_dist,
                            n = n,
                            effect_size = es,
                            F_p_value = mod$test(tmp_dat)$p_value,
                            BP_p_value = HETER_MODEL$test(tmp_dat)$p_value,
                            SW_p_value = shapiro.test(tmp_dat$.resid)$p.value,
                            boot_id = i)
                   }) %>%
  reduce(bind_rows)
  }

@@

                     es <- filter(vi_survey, 
                                  a == {{a}}, 
                                  b == {{b}},
                                  x_dist == {{x_dist}},
                                  n == {{n}}) %>%
                              head(1) %>%
                              pull(effect_size)

                     tibble(a = a,
                            b = b,
                            x_dist = x_dist,
                            n = n,
                            effect_size = es,
                            F_p_value = POLY_MODEL$test(mutate(tmp_dat, z = poly_model()$gen(n, computed = select(tmp_dat, x))$z))$p_value,
                            BP_p_value = mod$test(tmp_dat)$p_value,
                            SW_p_value = shapiro.test(tmp_dat$.resid)$p.value,
                            boot_id = i)
                   }) %>%
  reduce(bind_rows)
  }

@@

## P-value and power comparison

Visual inference and conventional hypothesis testing are two distinct statistical procedures. As displayed in Figure \ref{fig:p-value-comparison}, the visual test gives higher p-values than the conventional test for both polynomial and heteroskedasticity models in most of the time, suggesting that humans have lower confidence and much higher tolerance of the residual departures than the conventional test.

In particular, given 95% confidence level, both tests will reject the null hypothesis in only around 58% of the cases. And in 41% of the cases, the visual test does not reject the null hypothesis but the conventional test does. Since fail to reject the null hypothesis in a visual test means that there is no obvious visual discoveries found in the residual plot, analysts and the general public as the consumers of the output may not fully believe in the rejection of the null hypothesis even though the alternative hypothesis is true. Even if the rejection is accepted, the model violation may not be considered as impactful due to the fact that the departures are not clearly visible.

This makes it sounds like people are arrogantly ignoring the accuracy of conventional tests and trust too much to their visual perception, but it is not. In fact, the sensitivity of the conventional test could sometimes distract and discourage analysts from finding a simple but good linear approximation to the data. The rejection of the null hypothesis because of acceptable and negligible residual departure is not practically meaningful and useful. This also partly explains why abundance of literature suggesting the use of data plot in regression diagnostics.

```{r p-value-comparison, fig.height=4, fig.cap="Visual test p-value compared to conventional test p-value for lineups produced by heteroskedasticity and non-linearity model. The scatter plot is drawn on a square root scale. Every dot on the plot corresponds to a lineup. Most of the dots are fallen on the left of the y = x line indicating visual test p-value is almost always higher than conventional test p-value. The blue curve is a smoothing of the dots showing a positive trend. "}
vi_survey %>%
  filter(!null_lineup) %>%
  filter(!attention_check) %>%
  mutate(type = ifelse(type == "polynomial", "non-linearity", type)) %>%
  left_join(vi_survey %>% 
           count(unique_lineup_id) %>%
           rename(num_subject = n)) %>%
  group_by(unique_lineup_id) %>%
  summarise(across(everything(), first)) %>%
  ggplot() +
  geom_abline() +
  geom_point(aes(conventional_p_value, p_value), alpha = 0.4) +
  geom_smooth(aes(conventional_p_value, p_value), se = FALSE, span = 0.6) +
  scale_x_sqrt() +
  scale_y_sqrt() +
  facet_wrap(~type) +
  ylab("Visual test p-value on square root scale") +
  xlab("Conventional test p-value on square root scale") +
  theme_light()
```

Figure \ref{fig:power-poly-uniform} shows the estimated power of visual test on lineups with uniform distribution $X_{raw}$ against natural logarithm of the effect with comparison to the power of an exact test - F-test, and two other residual-based conventional tests commonly used in regression diagnostics but for testing other departures from the model assumptions. In overall, the power of all four tests increases as the effect becomes larger. The power curve of F-test climbs aggressively from 25% to around 90% as $log_e(E)$ increases from 0 to 2, while others respond inactively to the change of effect and remain lower than 25% throughout the period, showing that as an exact test, the F-test is relatively more sensitive to the type of model defects that being considered. The power of visual test arises steadily and nearly linearly to around 90% as $log_e(E)$ increases from 2 to 5, suggesting that the effect starts to make noticeable impact on the degree of the presence of the designed visual features. Other two inappropriate conventional tests shows improvement at the same time but at a lower rate. This coincides the point made by @cook1982residuals mentioned in \ref{hypothesis-testing} that residual-based tests for a specific type of model defect are sensitive to other types of model defects. At $log_e(E) \approx 6$, the power curve of F-test reaches almost 100% followed by the visual test by a small margin. The power of Breusch--Pagan test and Shapiro--Wilk test reach around 75% and 30% respectively.

@@

```{r eval = FALSE}
filter(vi_survey, type == "polynomial") %>%
  filter(x_dist == "uniform") %>%
  mutate(log_es = abs(log(effect_size) - 4)) %>%
  arrange(log_es) %>%
  View()
```

@@

```{r power-poly-uniform, fig.cap = "..."}
vi_survey %>%
  filter(x_dist == "uniform") %>%
  group_by(unique_lineup_id) %>%
  summarise(across(everything(), first)) %>%
  filter(type == "polynomial") %>%
  filter(e_sigma != 0.125) %>%
  mutate(reject_dirichlet = as.numeric(p_value <= 0.05)) %>%
  mutate(log_effect_size = log(effect_size)) %>%
  ggplot() +
  geom_smooth(aes(log_effect_size, reject_dirichlet, col = "Visual p-value"), method = "glm", method.args = list(family = binomial), se = FALSE) +
  stat_smooth(geom = "line", 
              data = ~boot_y(.x, reject_dirichlet, times = 500), 
              aes(log_effect_size, reject_dirichlet, group = boot_id, col = "Visual p-value"), 
              method = "glm", 
              method.args = list(family = binomial), 
              se = FALSE, 
              alpha = 0.03) +
  geom_smooth(data = mutate(poly_conv_sim, log_effect_size = log(effect_size)) %>%
                filter(x_dist == "uniform") %>%
                pivot_longer(F_p_value:SW_p_value) %>%
                mutate(name = gsub("_p_value", " p-value", name)), 
              aes(log_effect_size, as.numeric(value < 0.05), col = name), 
              method = "glm", 
              method.args = list(family = binomial),
              se = FALSE) +
  geom_point(aes(log_effect_size, prop_detect), alpha = 0.3) + geom_point(data = ~filter(.x, unique_lineup_id %in% c("poly_166", "poly_269", "poly_396", "poly_574")), aes(log_effect_size, prop_detect), alpha = 1, col = "red", size = 2) +
  theme_light(base_size = 5) +
  scale_color_brewer(palette = "Dark2", direction = -1) +
  xlab("Natural logarithm of effect size") +
  ylab("Power") -> p

VI_MODEL$plot_lineup(vi_lineup[["poly_166"]]$data, 
                     remove_grid_line = TRUE, 
                     theme = theme_light(base_size = 5),
                     remove_axis = TRUE,
                     size = 0.1) + 
  ggtitle(bquote(log[e](E) %~~% 1~","~"F-test p-value" == .(format(filter(vi_survey, unique_lineup_id == "poly_166")$conventional_p_value[1], digits = 3))~", Visual p-value " == .(format(filter(vi_survey, unique_lineup_id == "poly_166")$p_value[1], digits = 3)))) -> p1

VI_MODEL$plot_lineup(vi_lineup[["poly_269"]]$data,
                     remove_grid_line = TRUE, 
                     theme = theme_light(base_size = 5),
                     remove_axis = TRUE,
                     size = 0.1) + 
  ggtitle(bquote(log[e](E) %~~% 2~","~"F-test p-value" == .(format(filter(vi_survey, unique_lineup_id == "poly_269")$conventional_p_value[1], digits = 3))~", Visual p-value " == .(format(filter(vi_survey, unique_lineup_id == "poly_269")$p_value[1], digits = 3)))) -> p2

VI_MODEL$plot_lineup(vi_lineup[["poly_396"]]$data, 
                     remove_grid_line = TRUE, 
                     theme = theme_light(base_size = 5),
                     remove_axis = TRUE, 
                     size = 0.1) + 
  ggtitle(bquote(log[e](E) %~~% 3~","~"F-test p-value" == .(format(filter(vi_survey, unique_lineup_id == "poly_396")$conventional_p_value[1], digits = 3))~", Visual p-value " == .(format(filter(vi_survey, unique_lineup_id == "poly_396")$p_value[1], digits = 3)))) -> p3

VI_MODEL$plot_lineup(vi_lineup[["poly_574"]]$data, 
                     remove_grid_line = TRUE, 
                     theme = theme_light(base_size = 5),
                     remove_axis = TRUE,
                     size = 0.1) + 
  ggtitle(bquote(log[e](E) %~~% 4~","~"F-test p-value" == .(format(filter(vi_survey, unique_lineup_id == "poly_574")$conventional_p_value[1], digits = 3))~", Visual p-value " == .(format(filter(vi_survey, unique_lineup_id == "poly_574")$p_value[1], digits = 3)))) -> p4

### 2022-11-15 paper_comparison/paper_comparison/paper_comparison.tex
finish p-value cmp

Subjects recruited from Prolific received a fixed payment for
participating in the experiment. However, some subjects will try to
maximize their earnings for minimum effort. During the review of
submissions, if we found a subject objectively demonstrated clear
low-effort throughout the experiment, e.g., failed all attention checks,
we rejected the submission. The rejected submissions will be removed
immediately, and Prolific will automatically recruit another subject as
substitution. Therefore, we only paid for approved submissions and no
further data screening procedure needed to be applied on the collected
data.

In overall, there were a total of 8860 lineup evaluations made by 443
subjects in experiment I, experiment II and experiment III, where 886
lineup evaluations were attention checks and were not used in the
following analysis.

The collated dataset is provided in \texttt{vi\_survey} of the
\texttt{visage} \texttt{R} package.

\hypertarget{p-value-comparison}{%
\subsection{P-value comparison}\label{p-value-comparison}}

Visual inference and conventional hypothesis testing are two distinct
statistical procedures. As displayed in Figure
\ref{fig:p-value-comparison}, the visual test almost always gives higher
p-values than the conventional test in both polynomial and
heteroskedasticity models, suggesting that humans have lower confidence
and much higher tolerance of the residual departures than the numerical
procedure. In fact, only around 58\% of the cases, both tests will agree
on the rejection decision given 95\% confidence level. And in 41\% of
the cases, visual test does not reject the null hypothesis but
conventional test does. Although we know that the all the designed
lineups contain various degrees of residual departures and the
conventional test is telling truth, analysts and general public as the
consumers of the output will possibly not believe in or at least not
response to the ``false alarm'' if they can not see the departures
directly from the data plot and view it as impactful. The sensitivity of
the conventional test may distract and discourage us from finding a good
linear approximation to the data we try to model. This also partly
explains why abundance of literature suggesting the use of data plot in
regression diagnostics.

@@

\caption{...}\label{fig:p-value-comparison}
\end{figure}

\hypertarget{power-comparison}{%
\subsection{Power comparison}\label{power-comparison}}

Figure \ref{fig:power-poly-uniform} shows the estimated power of visual
test on lineups with uniform distribution \(X_{raw}\) against natural
logarithm of the effect with comparison to the power of an exact test -
F-test, and the power of two other residual-based conventional tests
commonly used in regression diagnostics but for testing other departures
from the model assumptions. In overall, the power of all four tests
increases as the effect becomes larger. The power curve of F-test climbs
aggressively from 25\% to around 90\% as \(log_e(E)\) increases from 0
to 2, while others respond inactively to the change of effect and remain
lower than 25\% throughout the period, showing that as an exact test,
the F-test is relatively more sensitive to the type of model defects
that being considered. The power of visual test arises steadily and
nearly linearly to around 90\% as \(log_e(E)\) increases from 2 to 5,
suggesting that the effect starts to make noticeable impact on the
degree of the presence of the designed visual features. Other two
inappropriate conventional tests shows improvement at the same time but
at a lower rate. This coincides the point made by
\citet{cook1982residuals} mentioned in \ref{hypothesis-testing} that
residual-based tests for a specific type of model defect are sensitive
to other types of model defects. At \(log_e(E) \approx 6\), the power
curve of F-test reaches almost 100\% followed by the visual test by a
small margin. The power of Breusch--Pagan test and Shapiro--Wilk test
reach around 75\% and 30\% respectively.

>>>>

There were 2880, 2880 and 2214 lineups evaluation made by 160, 160 and
123 subjects recruited for experiment I, II and III respectively. In the
total of 7974 lineup evaluations, 3744 used lineups produced by the
non-linearity model, 3510 used lineups produced by the
heteroskedasticity model, and 720 used lineups consists of only null
plots.

Besides, there were 886 attention checks not included in the following
analysis. The collated dataset is provided in \texttt{vi\_survey} of the
\texttt{visage} \texttt{R} package.

\hypertarget{p-value-and-power-comparison}{%
\subsection{P-value and power
comparison}\label{p-value-and-power-comparison}}

Visual inference and conventional hypothesis testing are two distinct
statistical procedures. As displayed in Figure
\ref{fig:p-value-comparison}, the visual test gives higher p-values than
the conventional test for both polynomial and heteroskedasticity models
in most of the time, suggesting that humans have lower confidence and
much higher tolerance of the residual departures than the conventional
test.

In particular, given 95\% confidence level, both tests will reject the
null hypothesis in only around 58\% of the cases. And in 41\% of the
cases, the visual test does not reject the null hypothesis but the
conventional test does. Since fail to reject the null hypothesis in a
visual test means that there is no obvious visual discoveries found in
the residual plot, analysts and the general public as the consumers of
the output may not fully believe in the rejection of the null hypothesis
even though the alternative hypothesis is true. Even if the rejection is
accepted, the model violation may not be considered as impactful due to
the fact that the departures are not clearly visible.

This makes it sounds like people are arrogantly ignoring the accuracy of
conventional tests and trust too much to their visual perception, but it
is not. In fact, the sensitivity of the conventional test could
sometimes distract and discourage analysts from finding a simple but
good linear approximation to the data. The rejection of the null
hypothesis because of acceptable and negligible residual departure is
not practically meaningful and useful. This also partly explains why
abundance of literature suggesting the use of data plot in regression
diagnostics.

@@

\caption{Visual test p-value compared to conventional test p-value for lineups produced by heteroskedasticity and non-linearity model. The scatter plot is drawn on a square root scale. Every dot on the plot corresponds to a lineup. Most of the dots are fallen on the left of the y = x line indicating visual test p-value is almost always higher than conventional test p-value. The blue curve is a smoothing of the dots showing a positive trend. }\label{fig:p-value-comparison}
\end{figure}

Figure \ref{fig:power-poly-uniform} shows the estimated power of visual
test on lineups with uniform distribution \(X_{raw}\) against natural
logarithm of the effect with comparison to the power of an exact test -
F-test, and two other residual-based conventional tests commonly used in
regression diagnostics but for testing other departures from the model
assumptions. In overall, the power of all four tests increases as the
effect becomes larger. The power curve of F-test climbs aggressively
from 25\% to around 90\% as \(log_e(E)\) increases from 0 to 2, while
others respond inactively to the change of effect and remain lower than
25\% throughout the period, showing that as an exact test, the F-test is
relatively more sensitive to the type of model defects that being
considered. The power of visual test arises steadily and nearly linearly
to around 90\% as \(log_e(E)\) increases from 2 to 5, suggesting that
the effect starts to make noticeable impact on the degree of the
presence of the designed visual features. Other two inappropriate
conventional tests shows improvement at the same time but at a lower
rate. This coincides the point made by \citet{cook1982residuals}
mentioned in \ref{hypothesis-testing} that residual-based tests for a
specific type of model defect are sensitive to other types of model
defects. At \(log_e(E) \approx 6\), the power curve of F-test reaches
almost 100\% followed by the visual test by a small margin. The power of
Breusch--Pagan test and Shapiro--Wilk test reach around 75\% and 30\%
respectively.
